---
title: "CEMiTool: Co-expression Modules Identification Tool"
output: 
  prettydoc::html_pretty:
    toc: true,
	theme: cayman,
	highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CEMiTool: Co-expression Modules Identification Tool}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Basic usage

The `CEMiTool` `R` package provides users with an easy-to-use method to automatically
implement `WGCNA` analyses with optimized parameters, gene set enrichment analysis and
overrepresentation analysis.

In order to use `CEMiTool`, you only needs a data.frame containing you expression data,
with the gene names in the rows and sample names in the columns, as following:

```{r}

library("CEMiTool")

# Load expression data
data(expr)
head(expr[,1:4])
```

There is a function called `cemitool` that receives the expression data and returns a `CEMiTool` object:
```{r}
cem <- cemitool(expr)
```

We can print the `CEMiTool` object to see what is inside of it:
```{r}
print(cem)
```

As you can see there are `r nmodules(cem)` and we the relation between genes and modules are in `cem@module`:
```{r}
head(cem@module)
```

As you include more informations to cemitool we can build a more complete object and as you will see a more reich report.

For instance, we can provide a sample annotation `data.frame` specifying a class for each sample.

```{r}
# Load your sample annotation data
data(sample_annotation_data)

head(sample_annotation_data)
```

And then generate a report:


```{r}
ncores <- parallel::detectCores()
doParallel::registerDoParallel(cores=ncores)
```

```{r}
# Reading GMT file
gmt_fname <- system.file("extdata", "pathways.gmt", package = "CEMiTool")
gmt_in <- read_gmt(gmt_fname)
```

```{r}
# Reading interactions
int_fname <- system.file("extdata", "interactions.tsv", package = "CEMiTool")
int_df <- read.delim(int_fname)
```

```{r}
# running cemitool
cem <- cemitool(expr, sample_annotation_data, gmt_in, interactions=int_df, 
                filter=F, plot=T, split_modules=T, sample_name_column="Sample")
```

```{r}
generate_report(cem)
```


```{r}
write_files(cem, directory="./reports", force=T)
```

```{r}
# create a new CEMiTool object
cem <- new("CEMiTool", expression=expr, 
               sample_annotation=sample_annotation_data,
               sample_name_column="Sample")
```

```{r}
# Find the modules
cem <- find_modules(cem) 
```

```{r}
# If you desire, split the modules by positive and negative correlation
cem <- split_modules(cem)
```

```{r}
# Take a look at the expression patterns of those modules.
# You can also use the gene_module variable here.
cem <- plot_profile(cem)
print(cem@profile_plot[[1]])
```

```{r}
# Now that you got the modules, why don't you just give a look
# at the enrichment of those modules in your experimental classes
cem <- mod_gsea(cem)
```

```{r}
# Heatmap of gene set enrichment analysis
cem <- plot_gsea(cem)
print(cem@enrichment_plot)
```

```{r}
# Performs over representation analysis
cem <- mod_ora(cem, gmt_in)
```

```{r}
# plot ora results
cem <- plot_ora(cem)
print(cem@barplot_ora[[1]][["pl"]])
```

```{r}
# adds interaction
cem <- include_interactions(cem, int_df)
```

```{r}
# plot interactions
cem <- plot_interactions(cem)
```

```{r results="hide"}
# generate report
generate_report(cem)
generate_report(cem, output_format=c("pdf_document", "html_document"))
```

